Computing in the Field of Rationals

### 2012-06-18 rational_ext.tex
_

\title{(With Extensions)The Two Dualities of Computation: Negative and Fractional Types}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), these types have particularly natural
  computational interpretations. Intuitively, values of negative types are
  values that flow ``backwards'' to satisfy demands and values of fractional
  types are values that impose constraints on their context.  The combination
  of these negative and fractional types enables greater flexibility in
  programming by breaking global invariants into local ones that can be
  autonomously satisfied by a subcomputation. Theoretically, these types give
  rise to \emph{two} function spaces and to \emph{two} notions of
  continuations, suggesting that the previously observed duality of
  computation conflated two orthogonal notions: an additive duality that
  corresponds to backtracking and a multiplicative duality that corresponds
  to constraint propagation.
\end{abstract}

@@

@@

>>>>

\title{Computing in the Field of Rationals}
%% \authorinfo{Roshan P. James}
%%            {Indiana University}
%%            {rpjames@indiana.edu}
%% \authorinfo{Amr Sabry}
%%            {Indiana University}
%%            {sabry@indiana.edu}
%% \maketitle

\begin{abstract}

  Previous work on \emph{information effects} established the notion
  of information preserving computation
  \cite{infeffects,rc2011,rc2012} and the core calculus for computing
  with the isomorphisms of finite types, {{langRev}}. Given the
  first-order nature of {{langRev}} however it was unclear how one may
  express first-class functions.  Andrej Filinski's masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations.

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systemmatically from the {{langRev}} by considering a symmetric
  notion of duality. Unlike the \lcal which shows a classical demorgan
  dulaity, here we have two axis of dualization and hence two
  dualities -- an additive duality, namely negative types and a
  multiplicative duality, namely fractional types.

  %% Every functional programmer knows about sum and product types, {{a+b}} and
  %% {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
  %% respectively, are much less known and their computational interpretation is
  %% unfamiliar and often complicated. We show that in a programming model in
  %% which information is preserved (such as the model introduced in our recent
  %% paper on \emph{Information Effects}), these types have particularly natural
  %% computational interpretations. Intuitively, values of negative types are
  %% values that flow ``backwards'' to satisfy demands and values of fractional
  %% types are values that impose constraints on their context.  The combination
  %% of these negative and fractional types enables greater flexibility in
  %% programming by breaking global invariants into local ones that can be
  %% autonomously satisfied by a subcomputation. Theoretically, these types give
  %% rise to \emph{two} function spaces and to \emph{two} notions of
  %% continuations, suggesting that the previously observed duality of
  %% computation conflated two orthogonal notions: an additive duality that
  %% corresponds to backtracking and a multiplicative duality that corresponds
  %% to constraint propagation.
\end{abstract}

@@

Andrej Filinksi's masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation is one embraces the symmetry hidden in
continuations. Filinski showed that values are dual to continuations,
functions are dual to delimited continuations. Filinksi also shows a
de-morgan style duality, by treating value pairs as the dual of
continuation sums and value sums as dual to continuation pairs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{First class functions from the permutation space.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Traces and the Int-Construction.}
